### 2025-03-13 paper/sections/experiments.qmd
main results finally done

$$
\text{implaus}_{\text{dist}}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
$$ {#eq-impl-dist}

@@

$$
\text{implaus}_{\text{div}}(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
$$ {#eq-impl-div}

@@

@@

***Hyperparameters for penalties.*** We find that the strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful; energy must be sufficiently regularized to avoid poor performance in terms of decreased plausibility and increased costs. The sensitivity with respect to $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$ is much less evident. While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes when combined with high values of $\lambda_{\text{div}}$ or $\lambda_{\text{adv}}$, this effect is not very pronounced.

***Other hyperparameters.*** We observe that the effectiveness and stability of CT is positively associated with the number of counterfactuals generated during each training epoch. We also confirm that a higher number of training epochs is beneficial. Interestingly, we find that it is not necessary to employ CT during the entire training phase to achieve the desired improvements in explainability. When training models conventionally during the first 50% of epochs before switching to CT for the next 50% of epochs, we observed positive results. Put differently, CT may be a way to improve the explainability of models in a fine-tuning manner.

>>>>

$$
\text{IP}(\mathbf{x}^\prime,\mathbf{X}^+) = \frac{1}{\lvert\mathbf{X}^+\rvert}\sum_{\mathbf{x} \in \mathbf{X}^+} \text{dist}(\mathbf{x}^{\prime},\mathbf{x})
$$ {#eq-impl-dist}

@@

$$
\text{IP}^*(\mathbf{X}^\prime,\mathbf{X}^+) = \text{MMD}(\mathbf{X}^\prime,\mathbf{X}^+)
$$ {#eq-impl-div}

@@

@tbl-main presents our main empirical findings. The top five rows show the percentage reduction in implausibility according to @eq-impl-dist for varying degrees of the energy penalty used for *ECCo* at test time. The following row shows the reduction in implausibility as measured by @eq-impl-div and aggregated across all test specifications of *ECCo*. The final two rows show the test accuracies for the model trained with CT and conventionally trained models ("vanilla").

We observe that for all datasets except *OL* and across all test settings, the average distance of counterfactuals from observed samples in the target class is reduced, indicating improved plausibility. The magnitude of improvements varies by dataset: for the simple synthetic datasets, distance reductions range from around 20-40% (*LS*, *Moon*) to almost 60% (*Circ*). For the real-world tabular datasets, improvements are generally smaller but still substantial in many cases with around 10-15% for *CH*, 11-28% for *GMSC*, 7-8% for *Cred* and around 3% for *Adult*. For our only vision dataset (*MNIST*), distances are reduced by up to 9%. The results for our proposed divergence metric are qualitatively similar, but generally even more pronounced: for the *Circ* dataset, implausibility is reduced by almost 94% to virtually zero as we verified by looking at the absolute outcome. Improvements for other datasets range from 28% (*Moon*) to 78% (*GMSC*). For *OL* the reduction is negative, consistent with the distance-based metric. The only dataset, for which our proposed metric disagrees with the distance-based metric is *MNIST*.

These broad and substantial improvements in plausibility generally do not come at the cost of decreased predictive performance: test accuracy for CT is virtually identical to the baseline for *Adult*, *Circ*, *LS*, *Moon* and *OL*, and even slightly improved for *Cred*. Exceptions to this general pattern are *MNIST*, *CH* and *GMSC*, for which we observe reduction in test accuracy of 2, 5 and 15 percentage points, respectively. We note in this context, that we have not optimized our models for predictive performance at all and worked with very small networks. In summary, we find that CT can substantially improve the quality of explanations learned by models without generally sacrificing predictive accuracy.

::: {#tbl-main}

```{=latex}
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}} cccccc }
\input{tables/main.tex}
\end{tabular*}
```

Key plausibility and predictive performance metrics for all datasets. The top five rows show the percentage reduction in implausibility according to @eq-impl-dist for varying degrees of the energy penalty used for *ECCo* at test time. The following row shows the reduction in implausibility as measured by @eq-impl-div and aggregated across all test specifications of *ECCo*. The final two rows show the test accuracies for the model trained with CT and conventionally trained models ("vanilla").

@@

***Hyperparameters for penalties.*** We find that the strength of the energy regularization, $\lambda_{\text{reg}}$ is highly impactful; energy must be sufficiently regularized to avoid poor performance in terms of decreased plausibility and increased costs. The sensitivity with respect to $\lambda_{\text{div}}$ and $\lambda_{\text{adv}}$ is much less evident. While high values of $\lambda_{\text{reg}}$ may increase the variability in outcomes when combined with high values for any of the other penalties in @eq-obj, this effect is not very pronounced.

***Other hyperparameters.*** We observe that the effectiveness and stability of CT is positively associated with the number of counterfactuals generated during each training epoch. We also confirm that a higher number of training epochs is beneficial. Interestingly, we find that it is not necessary to employ CT during the entire training phase to achieve the desired improvements in explainability. We have tested training models conventionally during the first half of training before switching to CT after this initial "burn-in" period and observed positive results. Put differently, CT may be a way to improve the explainability of trained models in a fine-tuning manner.

### 2025-03-13 paper/sections/other/main_results.qmd
main results finally done

df = final_table(res_dir)
get_table_inputs(df, nothing; backend=Val(:latex)) |>
    inputs -> tabulate_results(inputs; table_type=:tabular, save_name="paper/tables/main.tex")
```

>>>>

df = final_table(res_dir)
inputs = get_table_inputs(df, nothing; backend=Val(:latex))
tabulate_results(inputs; table_type=:tabular, save_name="paper/quarto_ecml/tables/main.tex", wrap_table=false)
```
